---
title: "calluna_xylem_cleaning"
format: html
editor: visual
---

# Calluna xylem cleaning

## Setup

```{r}
#| label: libraries
#| message: false
#| warning: false

library(readxl) # Pour lire Excel
library(writexl)  # Pour écrire Excel
library(dplyr)
library(openxlsx)
library(ggplot2)
library(tidyr)
library(lme4)
library(lmerTest)
library(glmmTMB)
library(ggeffects)
library(emmeans)
library(patchwork)
library(plotrix)
library(moments)
```

## Import results for 2024
Importing all files containing the results of the image processing for 2024 growth rings and combining them in one dataset

```{r}
#| label: Creating database with the results of 2024

# Import all the files containing the results for xylem area
xylem_area_files <- list.files("Data/results_xylem_area_2024_Loic", 
                    full.names = TRUE)

# Merging all file into one dataset
xylem_area_raw <- xylem_area_files |>
  setNames(tools::file_path_sans_ext(basename(xylem_area_files))) |>
  purrr::map_df(
    ~ readr::read_csv(.x, show_col_types = FALSE),
    .id = "source_file"
  ) |>
  rename(xylem_id = `...1`) # '...1' appear if a csv file first column doesn't have a name

# Extract information from the filename 
xylem_area_combined <- xylem_area_raw %>%
  separate_wider_delim(
    source_file,
    delim = "-",
    names = c(
      "age_class",
      "nin_rep",
      "direction",
      "RESULTS",
      "SUBSET",
      "subset"
    ),
    cols_remove = FALSE
  ) %>%
  mutate(
    plant = paste(age_class, nin_rep, sep = "-"),
    nin_type = stringr::str_extract(nin_rep, "^[A-Za-z]+"),
    treatment = paste(age_class, nin_type, sep = "-"),
    plant_direction = paste(age_class, nin_rep, direction, sep = "-"),
    subset_id = paste(age_class, nin_rep, direction, subset, sep = "-")
  ) %>%
  mutate(
    year = as.character("2024"),
    observer = as.character("LB"),
    age_class = factor(
      age_class, 
      levels = c("PIO", "BUI", "MAT")
      ),
    nin_type = factor(
      nin_type,
      levels = c("BH", "FH", "TH"))
  ) %>%
  relocate(
    age_class,
    .after = plant
  ) %>%
  relocate(
    subset,
    .after = plant_direction
  ) %>%
  select(
    -direction,
    -RESULTS,
    -SUBSET,
    -nin_rep
  )

write.csv(xylem_area_combined, "Output/results_2024.csv")
```

## Filter cut xylem
Because during the image processing all xylems were kept, a filtering of the xylems not fully inside the subset is necessary.

```{r}
#| label: Filtering of cut xylem

# Import  the file indicating the cut xylems
cut_partial_xylems <- read_excel("Data/cut_and_partial_xylems.xlsx")

# Adding cut and partial columns to the data
xylem_area_partial_combined <- xylem_area_combined %>%
  left_join(
    cut_partial_xylems  %>%
      select(subset_id, xylem_id, cut_xylem, partial_xylem),
    by = c("subset_id", "xylem_id")
  )

# Filtering only uncut xylems
xylem_area_cut_filtered <- xylem_area_partial_combined %>%
  filter(is.na(cut_xylem))

```

## Subsample area 2024
Import results for subset area

```{r}
#| label: Subsample area

# Import all files containing results for subsample area
files_subsample_area <- list.files("Data/subsample_area_2024", 
                    full.names = TRUE)

# Merging all file into one dataset
subset_area_raw <- files_subsample_area |>
  setNames(tools::file_path_sans_ext(basename(files_subsample_area))) |>
  purrr::map_df(
    ~ readr::read_csv(.x, show_col_types = FALSE),
    .id = "source_file"
  ) |>
  rename(subset = `...1`)

# Create subset_id identical to the column in the dataset of the xylem area results
subset_area_clean <- subset_area_raw %>%
  mutate(
    subset_id = paste(source_file, subset, sep = "-"),
    subset_area = Area
  ) 

# Functions to correct the order for the subset area
swap_subset_area <- function(data, id, subset1, subset2, subset3) {

  area_1 <- data$subset_area[
    data$source_file == id &
      data$subset == subset1
  ]
  
  area_2 <- data$subset_area[
    data$source_file == id &
      data$subset == subset2
  ]
  
  area_3 <- data$subset_area[
    data$source_file == id &
      data$subset == subset3
  ]

  data$subset_area[
    data$source_file == id &
      data$subset == 1
  ] <- area_1

  data$subset_area[
    data$source_file == id &
      data$subset == 2
  ] <- area_2
  
  data$subset_area[
    data$source_file == id &
      data$subset == 3
  ] <- area_3

  data
} # Place subset in the order established
swap_subset_area_2 <- function(data, id, subset1, subset2) {

  temp <- data$subset_area[
    data$source_file == id &
      data$subset == subset1
  ]

  data$subset_area[
    data$source_file == id &
      data$subset == subset1
  ] <- data$subset_area[
    data$source_file == id &
      data$subset == subset2
  ]

  data$subset_area[
    data$source_file == id &
      data$subset == subset2
  ] <- temp

  data
} # Swap to subset area together 

# Function application to obtain the correct order
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH1-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH1-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH4-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH5-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH5-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH7-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-BH7-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH1-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH1-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH2-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH2-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH3-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH4-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH4-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH5-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-FH5-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH3-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH3-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH5-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH5-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH6-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH6-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH7-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "BUI-TH7-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-BH1-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-BH2-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-BH2-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-BH3-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-FH1-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-FH2-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-FH3-N", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-FH5-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-TH1-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-TH2-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-TH3-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-TH5-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "MAT-TH5-S", 3, 2, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-BH1-N", 3, 1, 2)
subset_area_clean <- swap_subset_area_2(subset_area_clean, "PIO-BH2-N", 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-BH2-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-BH3-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-BH4-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-FH1-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-FH1-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-FH3-N", 3, 1, 2) 
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-FH3-S", 2, 3, 1)
subset_area_clean <- swap_subset_area_2(subset_area_clean, "PIO-TH1-N", 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH1-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH2-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH2-S", 1, 3, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH3-N", 3, 1, 2)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH3-S", 2, 3, 1)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH4-N", 2, 1, 3)
subset_area_clean <- swap_subset_area(subset_area_clean, "PIO-TH4-S", 2, 3, 1)

# Joining subsample area to xylem area
xylem_and_subset_data_raw <- xylem_area_cut_filtered %>%
  left_join(
    subset_area_clean  %>%
      select(subset_id, subset_area),
    by = c("subset_id")
  )

```

## Damage 2024

```{r}
#| label: Damage indices

xylem_damage <- read_excel("Data/damage_report_2024.xlsx")

xylem_damage <- xylem_damage %>%
  mutate(across(starts_with("damage_sub"), as.numeric))

damage_long <- xylem_damage %>%
  select(
    - subset1_missing,
    - subset2_missing,
    - subset3_missing
  ) %>%
  pivot_longer(
    cols = starts_with("damage_sub"),
    names_to = "subset",
    values_to = "subset_damage"
  ) %>%
  mutate(
    subset = sub("damage_sub", "", subset), # replace names like damage_sub1 by just 1
    whole_plant_damage = whole_plant_damage,
    cardinal_direction_damage = cardinal_direction_damage,
    warped = warped,
    folds = folds,
    late_season_missing = late_season_missing,
    plant_direction = image_id
  ) %>%
  relocate(
    subset_damage,
    .before = whole_plant_damage
  ) %>%
  select(-image_id)

raw_data <- xylem_and_subset_data_raw %>%
  left_join(
    damage_long,
    by = c("plant_direction", "subset")
  )

```

## Distribution of damage

```{r}
#| label: distribution of damage

damage_distribution4 <- raw_data %>%
  filter(subset_damage == 4) %>%
  count(treatment)

damage_distribution5 <- raw_data %>%
  filter(subset_damage == 5) %>%
  count(treatment)

sum_4 <- sum(damage_distribution4$n)
sum_5 <- sum(damage_distribution5$n)
```

## Evenness

```{r}
#| label: distribution of damage
#| message: false
#| warning: false

#this chunk creates a dataframe that count the amount of xylem per treatement before and after filtering by circularity

xylem_count_evenness <- raw_data %>%
  count(treatment)

xylem_count_evenness_filtered <- raw_data %>%
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) %>%
  filter(circ >= 0.65) %>%
  count(treatment)

xylem_count_combined <- xylem_count_evenness %>%
  mutate(count_filter = xylem_count_evenness_filtered$n) %>%
  mutate(percentage_remaining = count_filter/n)

names(xylem_count_combined) <- c("Treatment", 
                                 "Number of xylems", 
                                 "Number of xylems after filtering", 
                                 "Percentage xylem remaining")

```

## Circularity and damage

```{r}
#| label: circularity in relation to damage

#this chunk provides a comparison of datapoints before and after filtering by circularity >= 0.70.

data_circ <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

data_circ_filter <- data_circ |>
  filter(circ >= 0.70)

#creating new filtered datasets
xylem_damage_5 <- data_circ |>
  filter(subset_damage == 5)

xylem_damage_4 <- data_circ |>
  filter(subset_damage == 4)

xylem_damage_5_filtered <- data_circ_filter |>
  filter(subset_damage == 5)

xylem_damage_4_filtered <- data_circ_filter |>
  filter(subset_damage == 4)

#create a custom palette to use
palette_damage <- c("orange", "red")


plot_damage <- ggplot() + 
  geom_jitter(data = data_circ, 
              aes(x = treatment, y = Area), 
              color = "grey") + 
  geom_jitter(data = xylem_damage_4, 
              aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + 
  geom_jitter(data = xylem_damage_5, 
              aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + 
  geom_point(data = data_circ, 
             aes(x = treatment, y = means_treatment), 
             color = "blue") + 
  labs(title = "Unfiltered") + 
  ylab(label = "Xylem vessel area in µm") + 
  xlab(label = "NiN x Age class") + 
  theme(legend.position = "none") + 
  scale_color_manual(values = palette_damage) + 
  scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage_filtered <- ggplot() + 
  geom_jitter(data = data_circ, 
              aes(x = treatment, y = Area), color = "grey") + 
  geom_jitter(data = xylem_damage_4_filtered, 
              aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + 
  geom_jitter(data = xylem_damage_5_filtered, 
              aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + 
  geom_point(data = data_circ, 
             aes(x = treatment, y = means_treatment), color = "blue") + 
  labs(title = "Circularity >= 0.7") + 
  ylab(label = "Xylem vessel area in µm") + 
  xlab(label = "NiN x Age class") + 
  scale_color_manual(values = palette_damage) + 
  guides(color = guide_legend(title = "Subset damage index")) + 
  scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage + plot_damage_filtered
```
```{r}
#| label: damage vs circularity

damage_data <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  filter(subset_damage < 4)

circularity_plot <- ggplot(data = data_circ_filter) + 
  geom_point(aes(x = circ, y= Area), alpha = 0.4)

damage_plot <- ggplot(data = damage_data) + 
  geom_point(aes(x = circ, y= Area, color = subset_damage), alpha = 0.4)

circularity_plot + damage_plot

```

Need to check again

```{r}
#| label: potential problematic xylem

potential_problematic_xylem <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  group_by(source_file) |>
  mutate(subset_means = mean(Area)) |>
  mutate(problematic = case_when(
    circ <= 0.65 & Area >= (subset_means + (subset_means * 0.2)) ~ 1, #this basically checks for uncircular xylem that also are larger than the means, as stacked xylem should have a tendency to be larger than average.
    .default = 0
  )) |>
  filter(problematic == 0)

before_plot <- ggplot(data = data_circ) + 
  geom_point(aes(x = circ, y = Area), alpha = 0.4) + 
  facet_grid(cols = vars(age_class), rows = vars(nin_type))

after_plot <- ggplot(data = potential_problematic_xylem) + 
  geom_point(aes(x = circ, y = Area), alpha = 0.4) + 
  facet_grid(cols = vars(age_class), rows = vars(nin_type))

before_plot + after_plot
```
## Assessing skewness after filtering

```{r}
#| label: assessing skewness after filtering

data_skew55 <- data_circ |>
  filter(circ > 0.55)

data_skew60 <- data_circ |>
  filter(circ > 0.60)

data_skew65 <- data_circ |>
  filter(circ > 0.65)

data_skew70 <- data_circ |>
  filter(circ > 0.70)

skew_raw <- raw_data |>
  mutate(set = "raw")

skew_filtered55 <- data_skew55 |>
  mutate(set = "filter") 

skew_filtered60 <- data_skew60 |>
  mutate(set = "filter") 

skew_filtered65 <- data_skew65 |>
  mutate(set = "filter") 

skew_filtered70 <- data_skew70 |>
  mutate(set = "filter") 

skew_both55 <- bind_rows(skew_raw, skew_filtered55)

skew_both60 <- bind_rows(skew_raw, skew_filtered60)

skew_both65 <- bind_rows(skew_raw, skew_filtered65)

skew_both70 <- bind_rows(skew_raw, skew_filtered70)

skew_summary55 <- skew_both55 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary60 <- skew_both60 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary65 <- skew_both65 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary70 <- skew_both70 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

#skew_plot <- ggplot(data = skew_summary) + geom_point(aes(x = treatment, y = skew, color = set))

mean_plot55 <- ggplot(data = skew_summary55) + 
  geom_point(aes(x = treatment, y = mean, color = set)) + 
  geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + 
  ggtitle("Circularity > 0.55")

mean_plot60 <- ggplot(data = skew_summary60) + 
  geom_point(aes(x = treatment, y = mean, color = set)) + 
  geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + 
  ggtitle("Circularity > 0.60")

mean_plot65 <- ggplot(data = skew_summary65) + 
  geom_point(aes(x = treatment, y = mean, color = set)) + 
  geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + 
  ggtitle("Circularity > 0.65")

mean_plot70 <- ggplot(data = skew_summary70) + 
  geom_point(aes(x = treatment, y = mean, color = set)) + 
  geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + 
  ggtitle("Circularity > 0.70")

mean_plot55 + mean_plot60 + mean_plot65 + mean_plot70
```


```{r}
#| label: chi square test

data_chi <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  mutate(kept = case_when(
    circ < 0.70 ~ 0,
    circ >= 0.70 ~ 1
  )) |> 
  group_by(treatment) |>
  count(kept)

data_chi <- xtabs(n ~ treatment + kept, data = data_chi)

chisq.test(data_chi)

```
The chi squared test shows that the filtering rule affects each treatment differently, so it is important to quantify if this bias is acceptable or favors a particular treatment. 

```{r}

```

## Final cleaning

```{r}
#| label: adding the potential double xylem

cleaning_data <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.)))  

cleaning_data <- cleaning_data |>
    group_by(source_file) |>
    mutate(subset_means = mean(Area)) |>
    mutate(double_xylem = case_when(
    circ <= 0.65 & Area >= (subset_means + (subset_means * 0.2)) ~ 1, #this basically checks for uncircular xylem that also are larger than the means, as stacked xylem should have a tendency to be larger than average.
    .default = 0
  ))

```


```{r}
#| label: removing damage 4 and 5

#remove datapoints that have damage level 5
cleaning_data <- cleaning_data |>
  filter(subset_damage < 4)

```


```{r}
#| label: removing circularity < 0.60

cleaning_data <- cleaning_data |>
  filter(circ > 0.60)

```
 
### Export 

```{r}
write.csv(cleaning_data, "clean_dataset/DURIN_clean_LygraFire_field_trait_2024.csv")
```

## Exploration on partial xylem

```{r}
#| label: distribution of partial xylem
#| message: false
#| warning: false

#this chunk creates a dataframe that count the amount of xylem per treatement before and after filtering by partial xylem

xylem_count_partial <- raw_data %>%
  count(treatment)

xylem_area_cut_filtered <- xylem_area_partial_combined %>%
  filter(is.na(cut_xylem))

xylem_count_partial_filtered <- raw_data %>%
  filter(is.na(partial_xylem)) %>%
  count(treatment)

xylem_count_partial_and_filtered <- xylem_count_partial %>%
  mutate(count_filter = xylem_count_partial_filtered$n) %>%
  mutate(percentage_remaining = count_filter/n)

names(xylem_count_partial_and_filtered) <- c("Treatment", 
                                 "Number of xylems", 
                                 "Number of xylems after filtering", 
                                 "Percentage xylem remaining")
```

